---
title: "survival analysis"
finalized: 30 Jul 2026
author: India Schneider-Crease
---

this analysis evaluates whether recent evidence of taenia serialis infection is associated with the hazard of leader-tenure loss in male geladas. the workflow prepares monthly tenure and infection histories, fits and checks survival models, evaluates alternative infection classifications, and translates the primary model into predicted tenure and reproductive consequences.

```{r}
# read data
tenure = read.csv("~/tenure.csv", na.strings = c("", "NA", "N/A", "unknown"))
```

1. data preparation: this section checks the imported data, formats dates, documents missing values, removes males without known dates of birth, imputes missing social variables, and creates  derived variables used in later analyses.

```{r}
# load initial packages
library(dplyr)
library(tidyr)
library(lubridate)
library(mice)

# inspect the imported data and record the original sample size
glimpse(tenure)
tenure %>% summarise(n_rows = n(), n_males = n_distinct(code), n_tenures = n_distinct(tenure.id))

# format the date variables
date_vars = c("start_date_aligned", "end_date_aligned", "adjusted_end", "month", "cyst_date_aligned", "dob")
tenure = tenure %>% mutate(across(all_of(date_vars), mdy))

# confirm that the date variables were formatted correctly
tenure %>% select(all_of(date_vars)) %>% glimpse()

# identify all variables containing missing values before cleaning
tenure %>% summarise(across(everything(), ~ sum(is.na(.x)))) %>% pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>% filter(n_missing > 0) %>% arrange(desc(n_missing))

# identify males with unknown dates of birth
tenure %>% filter(is.na(dob)) %>% summarise(n_rows = n(), n_males = n_distinct(code))

# remove males with unknown dates of birth
tenure = tenure %>% filter(!is.na(dob))

# summarize
tenure %>% summarise(n_rows = n(), n_males = n_distinct(code), n_tenures = n_distinct(tenure.id))

# identify missing values in the social variables
to_impute = c("total_follower_males", "N.Females")
tenure %>% summarise(across(all_of(to_impute), ~ sum(is.na(.x))))
tenure %>% filter(if_any(all_of(to_impute), is.na)) %>% summarise(n_rows_with_missing_social_data = n())

# flag the values that will be imputed
tenure = tenure %>% mutate(across(all_of(to_impute), is.na, .names = "{.col}_imputed"))

# specify the variables included in the imputation model
imputation_vars = c("event", "stop", "cyst", "age_at_month", to_impute)
mice_method = make.method(tenure[imputation_vars])
mice_method[] = ""
mice_method[to_impute] = "pmm"

# impute the  social variables using predictive mean matching
imp = mice(tenure[imputation_vars], m = 1, maxit = 1, method = mice_method, seed = 1, printFlag = FALSE)

# extract the completed dataset
completed = complete(imp, 1)

# replace only the missing values with their imputed values
tenure = tenure %>% mutate(across(all_of(to_impute), ~ coalesce(.x, completed[[cur_column()]])))

# create binned social variables for modeling
tenure = tenure %>% mutate(group_size = factor(case_when(between(N.Females, 1, 4) ~ "small", between(N.Females, 5, 7) ~ "medium", between(N.Females, 8, 12) ~ "large", TRUE ~ NA_character_), levels = c("small", "medium", "large")))

# confirm that no missing values remain
tenure %>% summarise(across(all_of(to_impute), ~ sum(is.na(.x))))

# confirm that the derived social variables contain no missing values
tenure %>% summarise(across(c(all_of(to_impute), group_size, total_follower_males), ~ sum(is.na(.x))))

```

2. data summary and checks: this section summarizes the cleaned dataset and examines the distributions of tenure duration, cyst history, group size, and follower-male number.

```{r}
# summarize the cleaned df at the row level
tenure %>% summarise(n_rows = n(), n_males = n_distinct(code), n_tenures = n_distinct(tenure.id), n_observed_tenure_losses = n_distinct(tenure.id[event == 1]), n_right_censored_tenures = n_distinct(tenure.id) - n_distinct(tenure.id[event == 1]), n_cyst_positive_rows = sum(cyst == 1, na.rm = TRUE), n_cyst_negative_rows = sum(cyst == 0, na.rm = TRUE), n_cyst_missing_rows = sum(is.na(cyst)), n_cyst_positive_males = n_distinct(code[cyst == 1]), n_second_tenures = n_distinct(tenure.id[grepl("-2$", tenure.id)]))

# summarize cyst history at the male level
tenure %>% summarise(cyst_history = case_when(any(cyst == 1, na.rm = TRUE) ~ "ever cyst-positive", all(is.na(cyst)) ~ "no cyst observations", TRUE ~ "never cyst-positive"), .by = code) %>% count(cyst_history, name = "n_males")

# summarize group size and the number of follower males
tenure %>% count(group_size, name = "n_rows") %>% mutate(percent = round(100 * n_rows / sum(n_rows), 1))
tenure %>% count(total_follower_males, name = "n_rows")

# confirm that tenure length is constant within each tenure
tenure %>% summarise(n_tenure_lengths = n_distinct(tenure_length_aligned, na.rm = TRUE), .by = tenure.id) %>% filter(n_tenure_lengths > 1)

# create one row per tenure for repeated descriptive summaries
tenure_summary = tenure %>% mutate(cyst_history = if_else(any(cyst == 1, na.rm = TRUE), "ever cyst-positive", "never cyst-positive"), .by = code) %>% summarise(tenure_number = case_when(grepl("-1$", first(tenure.id)) ~ "first", grepl("-2$", first(tenure.id)) ~ "second", TRUE ~ "unclassified"), tenure_length_days = first(tenure_length_aligned), tenure_length_months = max(stop, na.rm = TRUE) - min(start, na.rm = TRUE), event_observed = any(event == 1, na.rm = TRUE), right_censored = !any(event == 1, na.rm = TRUE), .by = c(code, tenure.id, cyst_history))

# summarize number of events and right censorship
tenure_summary %>% count(event_observed, right_censored, name = "n_tenures") 

# summarize tenure duration overall
tenure_summary %>% summarise(n_tenures = n(), mean_tenure_days = mean(tenure_length_days, na.rm = TRUE), mean_tenure_years = mean(tenure_length_days, na.rm = TRUE) / 365.25, median_tenure_days = median(tenure_length_days, na.rm = TRUE), minimum_tenure_days = min(tenure_length_days, na.rm = TRUE), maximum_tenure_days = max(tenure_length_days, na.rm = TRUE))

# summarize first and second tenures
tenure_summary %>% filter(tenure_number != "unclassified") %>% group_by(tenure_number) %>% summarise(n_tenures = n(), mean_tenure_months = mean(tenure_length_months, na.rm = TRUE), median_tenure_months = median(tenure_length_months, na.rm = TRUE), minimum_tenure_months = min(tenure_length_months, na.rm = TRUE), maximum_tenure_months = max(tenure_length_months, na.rm = TRUE), .groups = "drop")

# inspect the distribution of observed tenure durations
hist(tenure_summary$tenure_length_days, main = "observed tenure durations", xlab = "tenure duration in days")

# summarize tenure duration by male cyst history
tenure_summary %>% group_by(cyst_history) %>% summarise(n_males = n_distinct(code), n_tenures = n(), mean_tenure_months = mean(tenure_length_months, na.rm = TRUE), median_tenure_months = median(tenure_length_months, na.rm = TRUE), minimum_tenure_months = min(tenure_length_months, na.rm = TRUE), maximum_tenure_months = max(tenure_length_months, na.rm = TRUE), .groups = "drop")

# summarize observed follow-up after the first cyst-positive month
tenure %>% filter(any(cyst == 1, na.rm = TRUE), .by = c(code, tenure.id)) %>% summarise(months_observed_after_cyst = max(stop, na.rm = TRUE) - min(start[cyst == 1], na.rm = TRUE), right_censored = any(end_follow_still_leader == 1, na.rm = TRUE), .by = c(code, tenure.id)) %>% summarise(n_tenures = n(), n_right_censored = sum(right_censored), mean_months_observed_after_cyst = mean(months_observed_after_cyst), minimum_months_observed_after_cyst = min(months_observed_after_cyst), maximum_months_observed_after_cyst = max(months_observed_after_cyst))
```

3. infection classification: this section links each monthly tenure observation to the most recent antigen result within the preceding 12 months. recent evidence of infection is defined as a positive antigen result within this window or a visible cyst during the current month. all other months are classified as having no recent evidence, including months with a recent negative result and months without a recent test.

```{r}
# read and format the antigen assay data
infection_data = read.csv("~/infection_data.csv", na.strings = c("", "NA", "N/A", "unknown")) %>% rename_with(tolower) %>% select(-any_of("x")) %>% mutate(collection_date = ymd(collection_date), positive = as.integer(positive))

# inspect assay data
glimpse(infection_data)
infection_data %>% summarise(n_samples = n(), n_males = n_distinct(code), n_missing_dates = sum(is.na(collection_date)), n_missing_results = sum(is.na(positive)))
infection_data %>% count(positive, name = "n_samples", .drop = FALSE)

# attach the most recent known assay result from the preceding 365 days
tenure = tenure %>%
  mutate(month_lower_bound = month - months(12)) %>%
  left_join(infection_data %>% select(code, collection_date, positive), by = join_by(code, closest(x$month >= y$collection_date), x$month_lower_bound <= y$collection_date), relationship = "many-to-one") %>%
  rename(last_sample_date = collection_date, last_sample_result = positive) %>%
  mutate(days_since_last_sample = as.integer(month - last_sample_date)) %>%
  select(-month_lower_bound)

# confirm that the join retains the original number of monthly observations
tenure %>% summarise(n_rows = n(), n_rows_with_recent_sample = sum(!is.na(last_sample_date)), minimum_days_since_sample = min(days_since_last_sample, na.rm = TRUE), maximum_days_since_sample = max(days_since_last_sample, na.rm = TRUE))

# inspect the time between the assay and the corresponding tenure month
hist(na.omit(tenure$days_since_last_sample), main = "time since most recent antigen test", xlab = "days")

# combine recent antigen results and current cyst observations
tenure = tenure %>% mutate(recent_infection_status = factor(case_when(cyst == 1 | last_sample_result == 1 ~ "detected", cyst == 0 & last_sample_result == 0 ~ "negative", TRUE ~ "unknown"), levels = c("unknown", "negative", "detected")),
  infection_evidence = factor(if_else(recent_infection_status == "detected", "detected", "not_detected"), levels = c("not_detected", "detected")))

# summarize the detailed and binary classifications
tenure %>% count(recent_infection_status, name = "n_rows")
tenure %>% count(infection_evidence, name = "n_rows")

# summarize infection evidence at observed tenure loss
tenure %>% filter(event == 1) %>% count(recent_infection_status, name = "n_tenure_losses")
tenure %>% filter(event == 1) %>% count(infection_evidence, name = "n_tenure_losses")

# create one record per tenure
tenure_level = tenure %>% summarise(duration_months = max(stop), event = max(event), .by = c(tenure.id, code))

# calculate descriptive observed durations
tenure_level %>% summarise(
  tenures = n(), males = n_distinct(code), tenure_losses = sum(event == 1, na.rm = TRUE),
  censored_tenures = sum(event == 0, na.rm = TRUE), mean_duration_months = mean(duration_months, na.rm = TRUE), median_duration_months = median(duration_months, na.rm = TRUE), minimum_duration_months = min(duration_months, na.rm = TRUE), maximum_duration_months = max(duration_months, na.rm = TRUE))

# plot the distribution of unique tenures
hist(tenure_level$duration_months, breaks = 15, main = "Observed tenure duration", xlab = "Tenure duration (months)")
```

4. primary model and model checks: this section evaluates support for the infection comparison within the event risk sets, fits the primary counting-process cox model, and checks the proportional-hazards and linearity assumptions. the model estimates tenure-loss hazard as a function of recent infection evidence, age, and follower-male number while allowing the baseline hazard to differ among group-size categories. the infection hazard ratio is also evaluated using a male-level bootstrap that resamples complete male histories.

```{r}
# load packages used for survival modeling and model summaries
library(survival)
library(broom)

# confirm that the model variables contain no missing values
tenure %>% summarise(across(c(start, stop, event, infection_evidence, age_at_month, total_follower_males, group_size, code, tenure.id), ~ sum(is.na(.x))))

# summarize the exposure distribution
tenure %>% summarise(n_rows = n(), n_males = n_distinct(code), n_tenures = n_distinct(tenure.id), n_events = sum(event), .by = infection_evidence)

# identify the event times within each group-size stratum
event_times = tenure %>% filter(event == 1) %>% mutate(event_time = stop) %>% summarise(n_events = n(), detected_events = sum(infection_evidence == "detected"), .by = c(event_time, group_size))

# examine whether both infection-evidence categories are represented in each event risk set
risk_set_support = event_times %>% left_join(tenure %>% select(tenure.id, start, stop, group_size, infection_evidence), by = join_by(group_size, x$event_time > y$start, x$event_time <= y$stop), relationship = "many-to-many") %>%
  summarise(n_events = first(n_events), detected_events = first(detected_events), at_risk_detected = n_distinct(tenure.id[infection_evidence == "detected"]), at_risk_not_detected = n_distinct(tenure.id[infection_evidence == "not_detected"]), .by = c(event_time, group_size)) %>% mutate(informative = at_risk_detected > 0 & at_risk_not_detected > 0)

# summarize support within the event risk sets
risk_set_support %>% summarise(n_event_risk_sets = n(), n_informative_risk_sets = sum(informative), n_events = sum(n_events), n_events_in_informative_risk_sets = sum(n_events[informative]), n_detected_events = sum(detected_events), n_detected_events_in_informative_risk_sets = sum(detected_events[informative]))

# fit the primary cox proportional hazards model
cm1 = coxph(Surv(start, stop, event) ~ infection_evidence + age_at_month + total_follower_males + strata(group_size), data = tenure, ties = "efron", id = tenure.id, cluster = code, x = TRUE)

# inspect the primary model
summary(cm1)

# create a numerical results table
coxresults_cm1 = tidy(cm1, exponentiate = TRUE, conf.int = TRUE) %>% select(term, hazard_ratio = estimate, conf_low = conf.low, conf_high = conf.high, p_value = p.value)
coxresults_cm1

# inspect model concordance
summary(cm1)$concordance

# test the proportional hazards assumption
ph_cm1 = cox.zph(cm1)
ph_cm1

# visually inspect the proportional hazards assumption
par(mfrow = c(1, 3))
plot(ph_cm1, var = "infection_evidence")
plot(ph_cm1, var = "age_at_month")
plot(ph_cm1, var = "total_follower_males")
par(mfrow = c(1, 1))

# add quadratic terms to check for nonlinear covariate effects
cm1_shape = update(cm1, . ~ . + I(age_at_month^2) + I(total_follower_males^2))

# inspect the quadratic terms
summary(cm1_shape)$coefficients[c("I(age_at_month^2)", "I(total_follower_males^2)"), , drop = FALSE]

# evaluate hazard ratio using male-level bootstrap that resamples complete male histories.

library(purrr)

# bootstrap the primary infection hazard ratio by resampling males
set.seed(123)
B = 5000
male_codes = unique(tenure$code)

bootstrap_hr = map_dfr(seq_len(B), function(b) {
  bootstrap_index = tibble(original_code = sample(male_codes, length(male_codes), replace = TRUE), bootstrap_copy = seq_along(original_code))

  bootstrap_data = bootstrap_index %>%
    inner_join(tenure %>% rename(original_code = code, original_tenure = tenure.id), by = "original_code", relationship = "many-to-many") %>%
    mutate(code = paste0(original_code, "_", bootstrap_copy), tenure.id = paste0(original_tenure, "_", bootstrap_copy))

  tryCatch({
    bootstrap_fit = suppressWarnings(coxph(Surv(start, stop, event) ~ infection_evidence + age_at_month + total_follower_males + strata(group_size), data = bootstrap_data, ties = "efron", id = tenure.id, cluster = code))
    infection_log_hr = unname(coef(bootstrap_fit)["infection_evidencedetected"])
    if (!is.finite(infection_log_hr)) stop("non-finite infection coefficient")
    tibble(rep = b, infection_log_hr, hazard_ratio = exp(infection_log_hr))
  }, error = function(e) NULL)
})

# inspect bootstrap completion 
bootstrap_hr %>% summarise(requested_bootstraps = B, successful_bootstraps = n(), percent_successful = round(100 * n() / B, 1),
    median_hazard_ratio = median(hazard_ratio), minimum_hazard_ratio = min(hazard_ratio), maximum_hazard_ratio = max(hazard_ratio))

# calculate the percentile bootstrap confidence interval
bootstrap_hr_ci = bootstrap_hr %>% summarise(method = "male-level percentile bootstrap", hazard_ratio = exp(unname(coef(cm1)["infection_evidencedetected"])), conf.low = exp(quantile(infection_log_hr, 0.025)), conf.high = exp(quantile(infection_log_hr, 0.975)))

# extract the model-based cluster-robust confidence interval
robust_hr_ci = tidy(cm1, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "infection_evidencedetected") %>%
  transmute(method = "model-based cluster-robust", hazard_ratio = estimate, conf.low, conf.high)

# compare the two confidence intervals
hr_ci_comparison = bind_rows(robust_hr_ci, bootstrap_hr_ci) %>% mutate(across(c(hazard_ratio, conf.low, conf.high), ~ round(.x, 2)))
hr_ci_comparison

# inspect the bootstrap distribution
hist(bootstrap_hr$infection_log_hr, breaks = 30, main = "bootstrap distribution of the infection coefficient", xlab = "log hazard ratio")
abline(v = unname(coef(cm1)["infection_evidencedetected"]), lwd = 2, lty = 2)

# create a results table with hazard ratios and percent changes in hazard
coxresults_cm1 = tidy(cm1, exponentiate = TRUE, conf.int = TRUE) %>%
  transmute(term,
    hazard_ratio = estimate, robust_conf_low = conf.low, robust_conf_high = conf.high, 
    percent_change = 100 * (estimate - 1), percent_conf_low = 100 * (conf.low - 1), percent_conf_high = 100 * (conf.high - 1),
    p_value = p.value) %>%
  left_join(bootstrap_hr_ci %>% transmute(term = "infection_evidencedetected",
        bootstrap_conf_low = conf.low, bootstrap_conf_high = conf.high, bootstrap_percent_conf_low = 100 * (conf.low - 1), bootstrap_percent_conf_high = 100 * (conf.high - 1)), by = "term") %>% mutate(predictor = case_match(term,
      "infection_evidencedetected" ~ "recent evidence of infection",
      "age_at_month" ~ "age",
      "total_follower_males" ~ "number of follower males",
      .default = term),
    hazard_ratio_ci = sprintf("%.2f (%.2f–%.2f)", hazard_ratio, robust_conf_low, robust_conf_high), percent_change_ci = sprintf("%.1f%% (95%% CI: %.1f%% to %.1f%%)", percent_change, percent_conf_low, percent_conf_high), bootstrap_ci = if_else(!is.na(bootstrap_conf_low), sprintf("%.2f–%.2f [%.1f%% to %.1f%%]", bootstrap_conf_low, bootstrap_conf_high, bootstrap_percent_conf_low, bootstrap_percent_conf_high), NA_character_), p_value = case_when(p_value < 0.001 ~ "<0.001", TRUE ~ sprintf("%.3f", p_value))) %>% select(predictor, hazard_ratio_ci, percent_change_ci, bootstrap_ci, p_value)

coxresults_cm1
```

5. six-month infection classification: this section links each monthly observation to the most recent prior antigen result and classifies recent evidence of infection using a narrower 6-month window.

```{r}
# attach the most recent test result without imposing the one-year limit
tenure = tenure %>% select(-any_of(c("last_test_date", "last_test_result", "last_test_status", "infection_history"))) %>% left_join(infection_data %>% select(code, last_test_date = collection_date, last_test_result = positive), by = join_by(code, closest(x$month >= y$last_test_date)), relationship = "many-to-one") %>% arrange(code, month)

# classify recent infection evidence using a six-month window
tenure = tenure %>% mutate(infection_evidence_6m = factor(if_else(coalesce(cyst == 1, FALSE) | coalesce(last_test_result == 1 & last_test_date >= month %m-% months(6), FALSE), "detected", "not_detected"), levels = c("not_detected", "detected")))

# examine rows and events under each classification
tenure %>% summarise(n_rows = n(), n_events = sum(event, na.rm = TRUE), .by = infection_evidence_6m)

# fit 6m model
cm1_6m = update(cm1, . ~ . - infection_evidence + infection_evidence_6m, data = tenure)
summary(cm1_6m)

# combine focal infection estimates across models
infection_results_table = bind_rows(
  tidy(cm1, exponentiate = TRUE, conf.int = TRUE) %>% mutate(model_order = 1, analysis = "primary", model = "12-month window"),
  tidy(cm1_6m, exponentiate = TRUE, conf.int = TRUE) %>% mutate(model_order = 2, analysis = "sensitivity", model = "6-month window")
) %>%
  filter(term %in% c("infection_evidencedetected", "infection_evidence_6mdetected")) %>%
  mutate(
    comparison = "detected vs not detected",
    hazard_ratio_95_ci = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
    percent_change_95_ci = sprintf("%.1f%% (%.1f%% to %.1f%%)", 100 * (estimate - 1), 100 * (conf.low - 1), 100 * (conf.high - 1)),
    p_value = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  arrange(model_order) %>%
  select(analysis, model, comparison, hazard_ratio_95_ci, percent_change_95_ci, p_value)
```


6. primary-model visualization: this section predicts adjusted tenure-retention curves for a male beginning tenure at the median starting age among males in medium groups with no follower males. follower number is fixed at zero, group size is fixed at medium, age increases monthly, and infection evidence is held constant across the prediction period.

```{r}
library(ggplot2)

# confirm factor levels used by the primary model
tenure = tenure %>% mutate(infection_evidence = factor(infection_evidence, levels = c("not_detected", "detected")), group_size = factor(group_size, levels = c("small", "medium", "large")))

# define the representative profile and plotting horizon
representative_group_size = "medium"
representative_followers = 0
representative_age = tenure %>% filter(start == 0, group_size == representative_group_size, total_follower_males == representative_followers) %>% summarise(value = median(age_at_month, na.rm = TRUE)) %>% pull(value)
plot_horizon = 70

# inspect the representative profile
tibble(starting_age = representative_age, total_follower_males = representative_followers, group_size = representative_group_size)

# create monthly covariate paths for both infection categories
curve_newdata = expand_grid(curve_id = factor(c("not_detected", "detected"), levels = c("not_detected", "detected")), start = 0:(plot_horizon - 1)) %>% mutate(stop = start + 1, event = 0, infection_evidence = factor(as.character(curve_id), levels = levels(tenure$infection_evidence)), age_at_month = representative_age + start / 12, total_follower_males = representative_followers, group_size = factor(representative_group_size, levels = levels(tenure$group_size)))

# predict adjusted survival curves and confidence intervals
conditional_survival = survfit(cm1, newdata = curve_newdata, id = curve_id, se.fit = TRUE, conf.type = "log-log")
curve_counts = as.integer(conditional_survival$strata)

# extract predicted survival estimates
survival_curves = tibble(time = conditional_survival$time, surv = as.numeric(conditional_survival$surv), lower = as.numeric(conditional_survival$lower), upper = as.numeric(conditional_survival$upper),
  group = rep(c("not_detected", "detected"), times = curve_counts)) %>%
  filter(time <= plot_horizon) %>% bind_rows(tibble(time = 0, surv = 1, lower = 1, upper = 1, group = c("not_detected", "detected"))) %>% arrange(group, time) %>% mutate(group = factor(if_else(group == "detected", "recent evidence of infection", "no recent evidence of infection"), levels = c("recent evidence of infection", "no recent evidence of infection")))

# create the primary model annotation
infection_label = tidy(cm1, exponentiate = TRUE, conf.int = TRUE) %>% filter(term == "infection_evidencedetected") %>%
  transmute(label = sprintf("adjusted HR = %.2f\n95%% CI: %.2f–%.2f\n%s", estimate, conf.low, conf.high, if_else(p.value < 0.001, "p < 0.001", sprintf("p = %.3f", p.value)))) %>% pull(label)

# define formatting shared by both panels
panel_theme = theme_classic(base_size = 22) + theme(plot.tag = element_text(size = 22))

survival_plot = ggplot(survival_curves, aes(time, surv, color = group)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = group), alpha = 0.18, color = NA, show.legend = FALSE) +
  geom_step(linewidth = 1.2) +
  annotate("text", x = Inf, y = Inf, label = infection_label, hjust = 1.08, vjust = 1.25, size = 8, lineheight = 1.15, color = "grey35") +
  scale_color_manual(values = c("recent evidence of infection" = "#B887B0", "no recent evidence of infection" = "#8CC9B8"), name = NULL) +
  scale_fill_manual(values = c("recent evidence of infection" = "#B887B0", "no recent evidence of infection" = "#8CC9B8"), name = NULL) +
  scale_x_continuous(breaks = c(0, 12, 24, 36, 48, 60), limits = c(0, 70), expand = expansion(mult = c(0, 0.01))) +
  scale_y_continuous(breaks = seq(0, 1, 0.25), limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 2.5))) +
  labs(x = "time (months)", y = "survival probability\n(tenure retained)", tag = "(a)") +
  panel_theme +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = "left", legend.box.just = "left",
  legend.text = element_text(size = 18), legend.background = element_rect(fill = "white", colour = "grey70", linewidth = 0.9))

survival_plot

# save to parallelize bootstrap run on HPC
saveRDS(tenure, "~/tenure_bootstrap_input.rds")
```

7. standardized tenure and reproductive predictions: this section reads the standardized rmst and bootstrap results, plots expected offspring under both infection-evidence conditions, and formats the rmst and expected-offspring estimates for reporting. the underlying analysis averages predictions across empirical tenure-start profiles while giving each male equal total weight, retaining each tenure’s starting age and group-size category, and fixing follower number at zero.

```{r}
# calculate rmst and bootstrap to calculate 95% CIs. see tenure_bootstrap.R for script run on ASU HPC Sol.

# read in rmst results
rmst_full = readRDS("~/rmst_full_results.rds")

# define the plotted time horizons
plot_taus = c(12, 24, 36, 48, 60)

# recover the mean observed tenure horizon from the results
average_tenure_horizon = rmst_full %>% filter(horizon == "mean observed tenure") %>% pull(tau)

# prepare expected-offspring estimates for plotting
offspring_plot_data = bind_rows(
  rmst_full %>% transmute(tau, offspring = offspring_not_detected, conf_low = offspring_not_detected_lwr, conf_high = offspring_not_detected_upr, group = "no recent evidence of infection"),
  rmst_full %>% transmute(tau, offspring = offspring_detected, conf_low = offspring_detected_lwr, conf_high = offspring_detected_upr, group = "recent evidence of infection")) %>%
  mutate(group = factor(group, levels = c("recent evidence of infection", "no recent evidence of infection")))

# plot expected offspring sired with the primary horizon and bootstrap confidence intervals
offspring_plot = ggplot(offspring_plot_data, aes(tau, offspring, color = group)) +
  geom_vline(xintercept = average_tenure_horizon, linetype = "dashed", linewidth = 1, color = "grey55") +
  geom_ribbon(aes(ymin = conf_low, ymax = conf_high, fill = group), alpha = 0.18, color = NA, show.legend = FALSE) +
  geom_line(linewidth = 1.2, show.legend = FALSE) +
  geom_point(size = 2, show.legend = FALSE) +
  annotate("text", x = average_tenure_horizon + 1, y = Inf, label = paste0("average tenure\n", round(average_tenure_horizon, 1), " months"), hjust = 0, vjust = 1.2, size = 8, color = "grey35") +
  scale_color_manual(values = c("recent evidence of infection" = "#B887B0", "no recent evidence of infection" = "#8CC9B8"), name = NULL) +
  scale_fill_manual(values = c("recent evidence of infection" = "#B887B0", "no recent evidence of infection" = "#8CC9B8"), name = NULL) +
  scale_x_continuous(breaks = plot_taus, labels = plot_taus, expand = expansion(mult = c(0, 0.01))) +
  scale_y_continuous(breaks = seq(0, 8, by = 2), labels = function(x) sprintf("%.2f", x), expand = expansion(mult = c(0, 0.02))) + coord_cartesian(ylim = c(0, 8)) + labs(x = "time horizon (months)", y = "expected offspring\nsired", tag = "(b)") + panel_theme + theme(legend.position = "none")

offspring_plot

# create the transposed rmst and expected-offspring table
rmst_results = bind_rows(
  rmst_full %>% transmute(row_order = 1, tau, measure = "RMST: no recent evidence detected, months", estimate_ci = sprintf("%.2f (%.2f to %.2f)", rmst_not_detected, rmst_not_detected_lwr, rmst_not_detected_upr)),
  rmst_full %>% transmute(row_order = 2, tau, measure = "RMST: recent evidence detected, months", estimate_ci = sprintf("%.2f (%.2f to %.2f)", rmst_detected, rmst_detected_lwr, rmst_detected_upr)),
  rmst_full %>% transmute(row_order = 3, tau, measure = "Months lost", estimate_ci = sprintf("%.2f (%.2f to %.2f)", lost_months, lost_months_lwr, lost_months_upr)),
  rmst_full %>% transmute(row_order = 4, tau, measure = "Expected offspring: no recent evidence detected", estimate_ci = sprintf("%.2f (%.2f to %.2f)", offspring_not_detected, offspring_not_detected_lwr, offspring_not_detected_upr)),
  rmst_full %>% transmute(row_order = 5, tau, measure = "Expected offspring: recent evidence detected", estimate_ci = sprintf("%.2f (%.2f to %.2f)", offspring_detected, offspring_detected_lwr, offspring_detected_upr)),
  rmst_full %>% transmute(row_order = 6, tau, measure = "Expected offspring lost", estimate_ci = sprintf("%.2f (%.2f to %.2f)", offspring_lost, offspring_lost_lwr, offspring_lost_upr))
) %>% mutate(tau = if_else(near(tau, round(tau)), paste0(sprintf("%.0f", tau), " months"), paste0(sprintf("%.1f", tau), " months"))) %>% pivot_wider(id_cols = c(row_order, measure), names_from = tau, values_from = estimate_ci) %>% arrange(row_order) %>% select(-row_order)

rmst_results
```

8. figure assembly: this section combines the adjusted tenure-retention and expected-offspring panels into a single figure with a shared infection-evidence legend.

```{r}
library(patchwork)

# combine panels with one shared legend
combined_figure = wrap_plots(survival_plot, guide_area(), offspring_plot,
  ncol = 1, heights = c(1, 0.12, 0.8), guides = "collect") 

combined_figure
```

